A Search Algorithm for the Global Extremum \break of a Discontinuous Function\thanks{This work was supported by the Russian Science Foundation, project No.\,21-11-00204.}

### 2021-10-10 2021/ICCS_BarkalovUsova/optima.tex
After answers for the reviewers

%%%% ДЛЯ РУССКОГО ТЕКСТА закомментировать потом!
\usepackage{inputenc}
\usepackage[english,russian]{babel}
\usepackage{cmap}
%%%%

@@

\begin{document}
%
\title{A search algorithm for the global extremum \break of a discontinuous function\thanks{This work was supported by the Russian Science Foundation, project No.\,21-11-00204.}}
%
\titlerunning{A search algorithm for the global extremum}
% If the paper title is too long for the running head, you can set 
% an abbreviated paper title here
%
\author{Konstantin Barkalov \orcidID{0000-0001-5273-2471}  
\and Marina Usova \orcidID{0000-0002-0722-6884}
}
%
\authorrunning{K. Barkalov, M. Usova}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhni Novgorod, Nizhni Novgorod, 
Russia \email{konstantin.barkalov@itmm.unn.ru}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
We consider the problem of finding the global minimum of a one-dimensional function that may have several finite jump discontinuity points. The discontinuities of the objective function can reflect the specifics of the optimized object in the mathematical model (for example, shock effects, resonance phenomena, jumps in geometric dimensions or material properties, etc.). In some cases, the discontinuing set is known. However, at the same time, there are problems in which there are no a priori estimates of discontinuity points, but it is known that such points are possible. The paper gives a description of the global search algorithm for solving the specified class of problems. The authors conducted an experimental comparison of the proposed algorithm with the methods implemented in MATLAB Global Optimization Toolbox.

@@

Global optimization problems involving many local extrema arise in many areas of applications (see, e.g., \cite{Cavoretto2021,Kvasov2013,Kvasov2015,Modorskii2016}). As a rule, the calculation of even a single value of the objective function in such problems requires considerable time, since the analytical form of the function is not known and the calculation of its values involves time-consuming numerical modelling. In this case, we have to deal with expensive black-box optimization problems as they are often called. Problems of this class are extremely complicated and are the object of intensive research.

One should note an important role of one-dimensional global optimization problems. Firstly, as has already been mentioned, they arise in many applications (see, e.g., \cite{Gillard2017,Sergeyev2020}). Secondly, many approaches to solving multidimensional optimization problems are, in one way or another, based on reducing the solution of the original multidimensional problem to solving a series of related one-dimensional optimization problems (see, for example, popular approaches in \cite{PaulaviciusZilinskas2014,Pinter1996,Sergeyev2017,Sergeyev2013}).

The literature describes various methods for solving the problem (\ref{problem}) depending on the assumptions about the properties of the objective function \cite{Horst1995,Horst1996,Pinter1996}. One of such assumptions, which is often fulfilled in applied optimization problems, is that the objective function $\varphi(x)$ satisfies the Lipschitz condition 
\[
\left|\varphi(x')-\varphi(x'')\right|\leq L\left|x'-x''\right|,\; x',x'' \in [a,b],
\]
where the constant $L$ is unknown a priori. 
This feature of the objective function is commonly used in many approaches to the development of algorithms for multiextremal optimization \cite{Evtushenko2009,Evtushenko2013,Jones2009,Paulavicius2011}.

However, in some applied problems, the Lipschitz property may not be fulfilled due to the presence of discontinuous jumps in the values of the function at some points of the search domain. Such jumps can be caused by sharp changes in the characteristics of the optimized object with small changes in its parameters, for example:
\begin{itemize}
  \item changes in the thickness of the material (or other geometric parameters of the object);
  \item changes in material properties (both when different materials are used in one design, and with changes in the properties of the same material as a result of technological operations, for example, welding, etc.);
  \item changes in external sources of impact (temperature, force, etc.).
\end{itemize}

We will interpret such sharp changes as jump discontinuities. For these cases, the set of points, at which the characteristics of the object jump, is typically known in advance. However, at the same time, there are problems in which there are no a priori estimates of discontinuity points, but it is known that such points are possible.

The known methods for solving such problems, as a rule, either generalize the concept of a gradient for discontinuous functions \cite{Batukhtin1993,Batukhtin1998,Moreau2000}, or belong to the class of bio-inspired algorithms \cite{Ban2019,ZhangXu}. These methods provide, generally speaking, the search for only a local solution to the problem.

In this paper, we will consider a deterministic algorithm for solving global optimization problems with both given and unspecified discontinuity points. This algorithm was developed on the basis of the \textit{global search algorithm} proposed by Strongin \cite{Strongin2000} and provides a search for the global optimum. 
%Russian
Проводимое в статье рассмотрение ограничено одномерным случаем. Обобщение нового алгоритма на многомерный случай может быть выполнено в соответствии со ставшими уже стандартными подходами из \cite{Grishagin2016,Grishagin2018}, использующими редукцию размерности на основе adaptive nested optimization scheme. 
Указанная схема редукции размерности позволяет заменить решение исходной многомерной задачи решением семейства одномерных задач, рекурсивно связанных между собой.

Основная часть статьи организована следующим образом.

Section 2 describes the algorithm for functions with given discontinuities. Section 3 discusses how to identify unspecified discontinuities. Section 4 describes how this discontinuity detection method can be used in the original algorithm. Section 5 presents the results of a comparison of the developed algorithm with the global search methods implemented in MATLAB Global Optimization Toolbox \cite{MatlabOTB}. Section~6 concludes the paper.

\section{Algorithm for functions with given discontinuity points}

Let us consider problem (\ref{problem}) assuming that the objective function $\varphi(x), x\in[a,b],$ has jump discontinuities at the given points
\begin{equation}\label{points}
a = \omega_0 <\omega_1 < ... <\omega_s < \omega_{s+1} = b
\end{equation}
and satisfies the Lipschitz condition between them, i.e.
\begin{equation}\label{LipschitzСondition}
\left|\varphi(x')-\varphi(x'')\right|\leq L\left|x'-x''\right|, \; x',x'' \in (\omega_{i-1},\omega_{i}), \; 1 \leq i \leq s+1.
\end{equation}
Points $\omega_{0}$ and $\omega_{s+1}$ are included in the general list regardless of the presence of function discontinuities in them for the convenience of subsequent presentation. Let us denote the resulting set of points $\Omega$, i.e. $\Omega = \{\omega_{0}, ..., \omega_{s+1}\}$, where $\omega_{0} = a, \; \omega_{s+1} = b$.

In \cite{Strongin2000}, the global search algorithm (GSA) was considered for solving problems without discontinuities. The authors also proposed a method for accounting for discontinuities based on a smoothing transformation of the objective function. Below we consider an algorithm based on GSA, in which the discontinuous function is minimized without smoothing of the discontinuities, which greatly simplifies the computational scheme of the algorithm.

@@

\item Check termination condition $x_t-x_{t-1} \leq \epsilon$, where $t$ is the number of the interval with the maximal characteristic and $\epsilon > 0$ is the predefined accuracy.

@@

%Russian
\textbf{Remark 1}. Еще раз отметим, что в соответствии с (\ref{gamma}) значение $\gamma_i+\gamma_{i-1}$ может быть равно либо 0 (обе граничных точки интервала принадлежат области непрерывности функции), либо 1 (одна из граничных точек является точкой разрыва). В первом случае формула (\ref{R_i}) будет иметь вид
\[
R(i)=\Delta_i+\frac{ (z_i-z_{i-1} )^2}{(r\mu)^2 \Delta_i }-\frac{2(z_i + z_{i-1})}{r\mu}.
\]
А во втором случае будем иметь
\[
R(i)=2\Delta_i - \frac{4 z_i }{r\mu}, \; R(i)=2\Delta_i - \frac{4 z_{i-1} }{r\mu},
\]
в случае разрыва в точке $x_{i-1}$ или $x_{i}$ соответственно.

\textbf{Remark 2}.
Приведенные выше формулы для вычисления характеристик интервалов могут быть использованы (в модифицированной форме) в алгоритме решения задач with non-convex constraints \cite{Barkalov2002,Sergeyev2006_1,Sergeyev2007,Sergeyev1995}. Здесь решение задачи with partially defined constraints неявно сводится к минимизации разрывной функции, в которой точками разрыва являются границы допустимого множества решений задачи.

@@

It can be proved that the conditions for the convergence of this algorithm will correspond to the conditions for the convergence of its basic prototype -- the global search algorithm from \cite{Strongin2000}. The formulation and rigorous proof of the convergence conditions will be the subject of further publications.

Let us illustrate the operation of the algorithm when searching for the minimum of the function
\begin{equation}\label{testFunction}
\varphi(x) =  0.1\sum_{i=1}^5{i\sin{(10(i+1)x+i)}}+
\begin{cases}
	-4, &0 \leq x < 3.1\\
	10, &3.1 \leq x < 4.6\\
	0, &4.6 \leq x < 7.0\\
	30, &7.0 \leq x < 9.0\\
	0, &9.0 \leq x \leq 10.0
\end{cases}
\end{equation}
at $x\in[0,10]$. All discontinuity points were considered known, the parameters of the method $r=2.0,\epsilon=0.001$ were used. Fig.~\ref{ris1} shows a graph of this function, the discontinuities are marked with a dotted line. The dashes under the graph indicate the points of 70 trials required for the algorithm to solve problem with the specified accuracy.

@@

Let us suppose now that the location of the discontinuity points for the function $\varphi(x)$ is not known, but it is known that such points are possible. Let the function have a discontinuity at an a priori unknown point $\omega \in(x_{i-1},x_i)$, then the relative difference $\mu_i$ from (\ref{mu_i}) will significantly exceed the relative differences for other intervals since within them the function satisfies the Lipschitz condition. Therefore, the identification of such situations can be based on a comparison of the relative differences $\mu_i$ from (\ref{mu_i}) corresponding to different search intervals $(x_{i-1},x_i)$.

@@

\begin{equation}\label{mu_cond}
\mu(1) \geq \mu(2) \geq ... \geq \mu(m)
\end{equation}
and define the minimum number $p$ satisfying the conditions
\begin{equation}\label{mu_main_cond}
 \frac{\mu(p)}{\mu(p+1)} \geq Q, \; 1 \leq p < q(m-2(s+1)),
\end{equation}
where $Q>1$ and $0<q<1$ are parameters.
%Russian
Значение параметра $Q$ определяет априорные предположения о том, во сколько раз относительная разность $\mu_i$, вычисленная на интервале $(x_{i-1},x_i)$, содержащем точку разрыва, будет превышать максимальную разность $\mu$, найденную по всем интервалам, принадлежащим области непрерывности функции. Данное значение задается исследователем, и в проведенных эксмериментах оно было выбрано равным 3.

When at the current $k$-th iteration of the method there is no $p$ that satisfies these conditions, it is interpreted as the absence of unspecified discontinuities. If such $p$ exists, this situation corresponds to the presence of undefined discontinuities. In this case, zero differences $\mu_i$ from (\ref{mu_i}), corresponding to intervals containing specified discontinuity points, are excluded from consideration by the condition $p<m-2(s+1)$.

Note that the value of the difference $\mu_i$ for the interval containing the extremum point can be close to zero, which will lead to the fulfillment of the first inequality from (\ref{mu_main_cond}). To eliminate the effect of small relative differences of intervals containing extreme points, the additional constraint $1 \leq p < q(m-2(s+1))$ is introduced in (\ref{mu_main_cond}) to exclude small values of $\mu_i$ from consideration. 
%Russian
Параметр $0<q<1$ тажке задается исследователем. Значение $1 - q$ отражает априорные предполжения о том, какая доля поисковых интервалов будет иметь значения $\mu_i$ близкими к нулю, т.е. фактически характеризует степень многоэкстремальности задачи. В проведенных нами экспериментах использовалось значение $q=0.9$.

@@

\section{Algorithm for unspecified discontinuities}

The algorithm is a modification of the algorithm described in Section 2 for given discontinuities and consists of the following.

@@

\begin{enumerate}
\item Sort points of a set $\{x^1,...,x^k\} \cup \Omega$ in ascending order of coordinates 
\begin{equation}\label{x_cond}
 a=x_0<x_1<...<x_{m-1}<x_m=b,
\end{equation}
where $m=s+k+1$. Each point is matched with an attribute $\gamma_i$ from (\ref{gamma}) (the presence or absence of a set discontinuity) and with the value of the function $z_i=\varphi(x_i)$ calculated only for $x_i \not\in \Omega$, i.e. at $\gamma_i=0$.

@@

\item Determine $p$ satisfying the inequalities (\ref{mu_main_cond}), where $0<q<1<Q$ are the parameters of the algorithm. Construct a set $I$ from (\ref{set_I}), the elements of which are interpreted as numbers of intervals that can contain a discontinuity. Generate interval attributes
\begin{equation}\label{delta_cond}
\delta_i=
\begin{cases}
	\sign{(z_i - z_{i-1})}, &\text{$i \in I$}, \\
	0, &\text{$i \not \in I$}.
\end{cases}
\end{equation}
The value $\delta_i=-1$ corresponds to a sharp decrease in the function in the interval $(x_{i-1},x_i)$, $\delta_i=1$ -- to an increase, $\delta_i=0$ -- to the lack of jump. Note that in accordance with the rules (\ref{gamma}) and (\ref{delta_cond}) for calculating attributes $\gamma_i$ and $\delta_i$ the inequalities
$ 0 \leq \gamma_i+\gamma_{i-1}+|\delta_i | \leq 1, \; 1 \leq i \leq m, $
are met.

@@

\item Check the termination condition $x_t-x_{t-1} \leq \epsilon$, where $t$ is the number of the interval with the maximal characteristic and $\epsilon > 0$ is the predefined accuracy.
\end{enumerate}

%Russian
\textbf{Remark}. В соответствии с (\ref{gamma}) and (\ref{delta_cond}) значение $\gamma_i+\gamma_{i-1}+|\delta_i |$ может быть равно 0 либо 1, в зависимости от наличия либо отсутствия точки разрыва в интервале $(x_{i-1},x_i)$.
Так, если данный интервал не содержит точек разрыва, т.е. $\gamma_i+\gamma_{i-1}+|\delta_i | = 0$, то его характеристика (\ref{R_i}) примет вид 
\[
R(i)=\Delta_i+\frac{ (z_i-z_{i-1} )^2}{(r\mu)^2 \Delta_i }-\frac{2(z_i + z_{i-1})}{r\mu}.
\]
Если одна из граничных точек интервала является заданной точкой разрыва, т.е. $\gamma_{i-1} = 1$ или $\gamma_i = 1$, то мы получим соответственно характеристики
\begin{equation}\label{RR}
R(i)=2\Delta_i - \frac{4 z_i }{r\mu}, \; R(i)=2\Delta_i - \frac{4 z_{i-1} }{r\mu}.
\end{equation}
Аналогичные формулы для характеристик интервалов будут получаться, если $(x_{i-1},x_i)$ является интервалом, который содержит незаданную точку разрыва, т.е. когда $|\delta_i | = 1 $. В этом случае мы таже будем получать формулы (\ref{RR}) для $\delta_i = -1$ и $\delta_i = 1$, что соответствует использованию результатов испытания в правой точке интервала для построения его характеристики в случае убывания функции и выбору левой точки при возрастании функции.

Fig.~\ref{ris2} shows the results of solving the problem from the previous example (\ref{testFunction}), in which the discontinuity points were considered unspecified. Here we used the parameters of the method $r=2.0, \epsilon=0.001, Q=3.0, q=0.3$. The dashes under the graph indicate the points of 70 trials required for the algorithm to solve problems with the specified accuracy.

@@

@@

The proposed algorithm also allows partial specification of known discontinuities. Fig.~\ref{ris3} shows the results of solving the same problem, for which the discontinuities were considered partially given (the discontinuities at the points 4.6 and 9.0 were considered to be known). Here we used the same parameters of the method $r=2.0, \epsilon=0.001, Q=3.0, q=0.3$. The dashes under the graph indicate the points of 67 trials required for the algorithm to solve problems with the specified accuracy.

@@

To demonstrate the efficiency of the global search algorithm for discontinuous functions (GSA-D) described in the previous section, we compare it with Direct Search (DS) \cite{Audet}, Simulated Annealing (SA) \cite{Kirkpatrick}, and Genetic Algorithm (GA) \cite{Goldberg} implemented in MATLAB Global Optimization Toolbox \cite{MatlabOTB}.

As the main comparison criterion, we will use the number of search trials $K$ (i.e., the number of calculations of the objective function) performed by the method before it converges.

The algorithms were compared when solving 1000 problems with a discontinuous objective function, which were constructed as follows. First, we generated continuous functions of the form
\begin{equation}\label{seriaFunction}
f(x) =  \sum_{j=0}^4{(j+1)A_j\sin{(5\pi jx+j)}+(j+1)B_j\cos(3\pi jx+j)}, x\in[0,1],
\end{equation}
where the values of the coefficients $A_j, B_j$ were chosen randomly and uniformly from the interval $[-6,6]$. Then 4 discontinuity points were added to the continuous functions. The coordinates of these points $\omega_1, \; \omega_2, \; \omega_3, \; \omega_4$ were set randomly and uniformly from the ranges $[0,0.25), \; [0.25,0.5), \; [0.5,0.75), \; [0.75,1]$ respectively. The jump values $\delta_1, \; \delta_2, \; \delta_3, \; \delta_4$ at discontinuities were also randomly selected from the range $[-50,50]$. These operations produced a discontinuous function of the form
\begin{equation}\label{seriaFunction_jump}
\varphi(x) =  f(x)+
\begin{cases}
	\delta_1, &0 \leq x < \omega_1\\
	\delta_2, &\omega_1 \leq x < \omega_2\\
	\delta_3, &\omega_2 \leq x < \omega_3 \;, \;\;\; x\in[0,1].\\
	\delta_4, &\omega_3 \leq x < \omega_4\\
	\delta_1, &\omega_4 \leq x \leq 1.0
\end{cases}
\end{equation}

Since the solution of the test problem assumes that the global optimizer $x^*$ is known, its value was preemptively estimated for each function of the series by iterating through all nodes of a uniform grid, the step of which was chosen to be sufficiently small.

>>>>

%%%% ДЛЯ РУССКОГО ТЕКСТА закомментировать потом!
%\usepackage{inputenc}
%\usepackage[english,russian]{babel}
%\usepackage{cmap}
%%%%

@@

\begin{document}
%
\title{A Search Algorithm for the Global Extremum \break of a Discontinuous Function\thanks{This work was supported by the Russian Science Foundation, project No.\,21-11-00204.}}
%
\titlerunning{A Search Algorithm for the Global Extremum}
% If the paper title is too long for the running head, you can set 
% an abbreviated paper title here
%
\author{Konstantin Barkalov \orcidID{0000-0001-5273-2471}  
\and Marina Usova \orcidID{0000-0002-0722-6884}
}
%
\authorrunning{K. Barkalov, M. Usova}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Lobachevsky State University of Nizhny Novgorod, Nizhny Novgorod, 
Russia \email{konstantin.barkalov@itmm.unn.ru}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
This article discusses the problem of finding the global minimum of a one-dimensional function that may have several finite jump discontinuity points. The discontinuities of the objective function may be due to the nature of the optimized object in the mathematical model (for example, shock effects, resonance phenomena, jumps in geometric dimensions or material properties, etc.). In some cases, the discontinuity poits are known; but at the same time, there are problems with no a priori estimates of discontinuity points. It is known, however, that such points do exist. The paper gives a description of the global search algorithm for solving such class of problems. In addition, the authors conducted an experimental comparison of the proposed algorithm with the methods of MATLAB Global Optimization Toolbox.

@@

Global optimization problems involving many local extrema arise in many applications (see, e.g. \cite{Cavoretto2021,Kvasov2013,Kvasov2015,Modorskii2016}). As a rule, the calculation of even a single value of the objective function in these problems takes considerable time, since the analytical form of the function is not known and the calculation of its values involves time-consuming numerical modelling. That is why these computationally expensive optimization problems are often called black-box problems. They are extremely complicated and are the subject of intensive research.

Despite their complex nature, these problems play an important role in many aspects of mathematical research. First, as has already been mentioned, they are applied for multiple tasks in a variety of fields (see, e.g., \cite{Gillard2017,Sergeyev2020}). Secondly, many approaches to solving multidimensional optimization problems, in one way or another, are based on reducing the solution of the original multidimensional problem to solving a series of related one-dimensional optimization problems (see, for example, popular approaches in \cite{PaulaviciusZilinskas2014,Pinter1996,Sergeyev2017,Sergeyev2013}).

The literature describes various methods for solving the problem (\ref{problem}) depending on the assumptions about the properties of the objective function \cite{Horst1995,Horst1996,Pinter1996}. One of such assumptions, which is often made about  applied optimization problems, is that the objective function $\varphi(x)$ satisfies the Lipschitz condition 
\[
\left|\varphi(x')-\varphi(x'')\right|\leq L\left|x'-x''\right|,\; x',x'' \in [a,b],
\]
where the constant $L$ is unknown a priori. 
This feature of the objective function is used for the development of algorithms for multiextremal optimization \cite{Evtushenko2009,Evtushenko2013,Jones2009,Paulavicius2011}.

However, in some applied problems, the Lipschitz condition may not be satisfied due to the presence of discontinuous jumps in the values of the function at certain points of the search domain. Such jumps can be caused by sharp changes in the characteristics of the optimized object which trigger minor changes in its parameters, for example:
\begin{itemize}
  \item changes in the thickness of material (or other geometric parameters of the object);
  \item changes in physical properties (both when different materials are used in a single design project, and when the properties of one material change as a result of technological operations, for example,  welding, etc.);
  \item changes in external sources of impact (temperature, force, etc.).
\end{itemize}

We will interpret such sharp changes as jump discontinuities. For these cases, the set of points, at which the characteristics of the object jump, is typically known in advance. However, there are problems with no a priori estimates of discontinuity points. It is known, however, that such points may exist.

The known methods for solving such problems, as a rule, either generalize the concept of a gradient for discontinuous functions \cite{Batukhtin1993,Batukhtin1998,Moreau2000}, or belong to the class of bio-inspired algorithms \cite{Ban2019,ZhangXu}. Generally speaking, these methods only search for a local solution to the problem.

In this paper, we will consider a deterministic algorithm for solving global optimization problems with both given and unspecified discontinuity points. This algorithm was developed on the basis of the \textit{global search algorithm} proposed by Strongin \cite{Strongin2000}, and it can be used to find the global optimum. 
%Russian
The scope of this article is limited to the one-dimensional case. Generalization of the new algorithm to the multidimensional case can be performed in accordance with the already existing standard approaches from \cite{Grishagin2016,Grishagin2018}, which use a dimensionality reduction based on the adaptive nested optimization scheme. 
The indicated dimensionality reduction scheme allows replacing the solution of the original multidimensional problem with the solution of a family of one-dimensional problems that are recursively related to each other.

The main part of the article is organized as follows.
Section 2 describes the algorithm for functions with given discontinuities. Section 3 discusses how to identify unspecified discontinuities. Section 4 describes how the discontinuity detection method can be used in the original algorithm. Section 5 presents the results of a comparison of the developed algorithm with the global search methods implemented in MATLAB Global Optimization Toolbox \cite{MatlabOTB}. Section~6 concludes the paper.

\section{Algorithm for Functions with Given Discontinuity Points}

Let us consider the problem (\ref{problem}) assuming that the objective function $\varphi(x), x\in[a,b],$ has jump discontinuities at the given points
\begin{equation}\label{points}
a = \omega_0 <\omega_1 < ... <\omega_s < \omega_{s+1} = b
\end{equation}
and satisfies the Lipschitz condition between these points, i.e.
\begin{equation}\label{LipschitzСondition}
\left|\varphi(x')-\varphi(x'')\right|\leq L\left|x'-x''\right|, \; x',x'' \in (\omega_{i-1},\omega_{i}), \; 1 \leq i \leq s+1.
\end{equation}
Points $\omega_{0}$ and $\omega_{s+1}$ are included in the general list regardless of the presence of function discontinuities at them for the convenience of subsequent presentation. Let us denote the resulting set of points $\Omega$, i.e. $\Omega = \{\omega_{0}, ..., \omega_{s+1}\}$, where $\omega_{0} = a, \; \omega_{s+1} = b$.

In \cite{Strongin2000}, the global search algorithm (GSA) was applied for solving problems without discontinuities. The authors also proposed a method for accounting for discontinuities based on a smoothing transformation of the objective function. Below we consider an algorithm based on GSA, in which the discontinuous function is minimized without smoothing of the discontinuities, which substantially simplifies the computational scheme of the algorithm.

@@

\item Check the stopping condition $x_t-x_{t-1} \leq \epsilon$, where $t$ is the number of the interval with the maximal characteristic and $\epsilon > 0$ is the predefined accuracy.

@@

%Russian
\textbf{Remark 1}. In accordance with (\ref{gamma}), the value $\gamma_i+\gamma_{i-1}$ can be equal to either 0 (both boundary points of the interval belong to the domain of continuity of the function), or 1 (one of the boundary points is the jump point). In the first case, the formula  (\ref{R_i}) will have the form
\[
R(i)=\Delta_i+\frac{ (z_i-z_{i-1} )^2}{(r\mu)^2 \Delta_i }-\frac{2(z_i + z_{i-1})}{r\mu}.
\]
And in the second case we will have
\[
R(i)=2\Delta_i - \frac{4 z_i }{r\mu}, \; R(i)=2\Delta_i - \frac{4 z_{i-1} }{r\mu},
\]
in case of discontinuity at the point  $x_{i-1}$ or $x_{i}$ respectively.

\textbf{Remark 2}.
The above formulas for calculating the characteristics of intervals can be used (in a modified form) in the algorithm for solving problems with non-convex constraints  \cite{Barkalov2002,Sergeyev2006_1,Sergeyev2007,Sergeyev1995}. Here, the solution to the problem with partially defined constraints is implicitly reduced to minimizing a discontinuous function, in which the discontinuity points are the boundaries of the admissible set of the constrained problem.

@@

It can be proved that the convergence conditions for this algorithm correspond to the convergence conditions for its prototype -- the global search algorithm from \cite{Strongin2000}. The formulation and rigorous proof of the convergence conditions will be the subject of further publications.

Let us illustrate the operation of the algorithm during the search for the minimum of the function
\begin{equation}\label{testFunction}
\varphi(x) =  0.1\sum_{i=1}^5{i\sin{(10(i+1)x+i)}}+
\begin{cases}
	-4, &0 \leq x < 3.1\\
	10, &3.1 \leq x < 4.6\\
	0, &4.6 \leq x < 7.0\\
	30, &7.0 \leq x < 9.0\\
	0, &9.0 \leq x \leq 10.0
\end{cases}
\end{equation}
at $x\in[0,10]$. Here, all the discontinuity points were considered to be known, and the parameters of the method were $r=2.0,\; \epsilon=0.001$. Fig.~\ref{ris1} shows a graph of this function; the discontinuities are marked with a dotted line. The dashes under the graph indicate the points of 70 trials required for the algorithm to solve problem with the specified accuracy.

@@

Let us suppose that the location of the discontinuity points for the function $\varphi(x)$ is not known, but it is known that such points are possible. Let the function have a discontinuity at an a priori unknown point $\omega \in(x_{i-1},x_i)$, then the relative difference $\mu_i$ from (\ref{mu_i}) will significantly exceed the relative differences for other intervals, since within within these intervals the function satisfies the Lipschitz condition. Therefore, such instances can be identified based on a comparison of the relative differences $\mu_i$ from (\ref{mu_i}) corresponding to different search intervals $(x_{i-1},x_i)$.

@@

\begin{equation}\label{mu_cond}
\mu(1) \geq \mu(2) \geq ... \geq \mu(m)
\end{equation}
and define the minimum number $p$ satisfying the conditions
\begin{equation}\label{mu_main_cond}
 \frac{\mu(p)}{\mu(p+1)} \geq Q, \; 1 \leq p < q(m-2(s+1)),
\end{equation}
where $Q>1$ and $0<q<1$ are parameters.
%Russian
The value of the parameter $Q$ determines the a priori assumptions about how many times the relative difference $\mu_i$, calculated on the interval $(x_{i-1},x_i)$ with the jump point, will exceed the maximum difference $\mu$, found over all intervals belonging to the domain of continuity of the function. This value is set by the researcher. In our experiments it was equal to 3.

The lack of $p$ at the $k$-th iteration of the method that satisfies these conditions is interpreted as the absence of unspecified discontinuities. If such $p$ exists, it indicates the existence of undefined discontinuities. In this case, zero differences $\mu_i$ from (\ref{mu_i}) corresponding to intervals containing specified discontinuity points, are excluded from consideration by the condition $p<m-2(s+1)$.

Note that the value of the difference $\mu_i$ for the interval containing the extremum point can be close to zero, which will lead to the fulfillment of the first inequality from (\ref{mu_main_cond}). To eliminate the effect of small relative differences of intervals containing extreme points, the additional constraint $1 \leq p < q(m-2(s+1))$ is introduced in (\ref{mu_main_cond}) to exclude small values of $\mu_i$ from consideration. 
%Russian
The parameter $0<q<1$ is also set by the researcher. The value $1 - q$ reflects a priori assumptions about what proportion of search intervals will have $\mu_i$ values close to zero, i.e. actually characterizes the degree of multi-extremity of the problem. In our experiments, we used the value $q=0.3$.

@@

\section{Algorithm for Unspecified Discontinuities}

This algorithm is a modification of the algorithm described in Section 2 for given discontinuities and consists of the following.

@@

\begin{enumerate}
\item Sort points of a set $\{x^1,...,x^k\} \cup \Omega$ in ascending order by their coordinates 
\begin{equation}\label{x_cond}
 a=x_0<x_1<...<x_{m-1}<x_m=b,
\end{equation}
where $m=s+k+1$. Each point is matched with an attribute $\gamma_i$ from (\ref{gamma}) (the presence or absence of a given discontinuity) and with the value of the function $z_i=\varphi(x_i)$ calculated only for $x_i \not\in \Omega$, i.e. at $\gamma_i=0$.

@@

\item Determine $p$ which satisfies the inequalities (\ref{mu_main_cond}), where $0<q<1<Q$ are the parameters of the algorithm. Construct a set $I$ from (\ref{set_I}), the elements of which are interpreted as numbers of intervals that can contain a discontinuity. Generate interval attributes
\begin{equation}\label{delta_cond}
\delta_i=
\begin{cases}
	\sign{(z_i - z_{i-1})}, &\text{$i \in I$}, \\
	0, &\text{$i \not \in I$}.
\end{cases}
\end{equation}
The value $\delta_i=-1$ corresponds to a sharp decrease in the function in the interval $(x_{i-1},x_i)$, $\delta_i=1$ -- to an increase, $\delta_i=0$ -- to the lack of jump. Note that in accordance with the rules (\ref{gamma}) and (\ref{delta_cond}) for calculating attributes $\gamma_i$ and $\delta_i$ the inequalities
$ 0 \leq \gamma_i+\gamma_{i-1}+|\delta_i | \leq 1, \; 1 \leq i \leq m, $
are satisfied.

@@

\item Check the stopping condition $x_t-x_{t-1} \leq \epsilon$, where $t$ is the number of the interval with the maximal characteristic and $\epsilon > 0$ is the predefined accuracy.
\end{enumerate}

%Russian
\textbf{Remark}. According to (\ref{gamma}) and (\ref{delta_cond}) the value $\gamma_i+\gamma_{i-1}+|\delta_i |$ can be 0 or 1, depending on the presence or absence of a jump in the interval  $(x_{i-1},x_i)$.
So, if the given interval does not contain jumps, i.e. $\gamma_i+\gamma_{i-1}+|\delta_i | = 0$, then its characteristic  (\ref{R_i}) takes the form
\[
R(i)=\Delta_i+\frac{ (z_i-z_{i-1} )^2}{(r\mu)^2 \Delta_i }-\frac{2(z_i + z_{i-1})}{r\mu}.
\]
If one of the boundary points of the interval is a given discontinuity, i.e. $\gamma_{i-1} = 1$ or $\gamma_i = 1$, then we get the characteristics
\begin{equation}\label{RR}
R(i)=2\Delta_i - \frac{4 z_i }{r\mu}, \; R(i)=2\Delta_i - \frac{4 z_{i-1} }{r\mu}.
\end{equation}
Similar formulas for the characteristics of intervals will be obtained if $(x_{i-1},x_i)$ is an interval that contains an unspecified discontinuity, i.e. when $|\delta_i | = 1 $. In this case, we will also obtain formulas (\ref{RR}) for $\delta_i = -1$ and $\delta_i = 1$, which correspond to using the trial result at the right point of the interval to compute its characteristic in the case of decreasing function and choosing left point as the function increases.

@@

Fig.~\ref{ris2} shows the results of solving the problem from the previous example (\ref{testFunction}), in which the discontinuity points were unspecified. Here we used the parameters of the method $r=2.0, \epsilon=0.001, Q=3.0, q=0.3$. The dashes under the graph indicate the points of 70 trials required for the algorithm to solve problems with the specified accuracy.

@@

The proposed algorithm also allows for partial specification of known discontinuities. Fig.~\ref{ris3} shows the results of solving the same problem, for which the discontinuities were considered partially given (the discontinuities at the points 4.6 and 9.0 were considered to be known). Here we used the same parameters of the method $r=2.0, \epsilon=0.001, Q=3.0, q=0.3$. The dashes under the graph indicate the points of 67 trials required for the algorithm to solve problems with the specified accuracy.

@@

To demonstrate the efficiency of the global search algorithm for discontinuous functions (GSA-D) described in the previous section, we compared it with Direct Search (DS) \cite{Audet}, Simulated Annealing (SA) \cite{Kirkpatrick}, and Genetic Algorithm (GA) \cite{Goldberg} implemented in MATLAB Global Optimization Toolbox \cite{MatlabOTB}.

We  used the number of search trials $K$ (i.e., the number of calculations of the objective function) performed by the method before it converges as the main comparison criterion.

The algorithms were compared on 1000 problems with a discontinuous objective function, which were constructed as follows. First, we generated continuous functions of the form
\begin{equation}\label{seriaFunction}
f(x) =  \sum_{j=0}^4{(j+1)A_j\sin{(5\pi jx+j)}+(j+1)B_j\cos(3\pi jx+j)}, x\in[0,1],
\end{equation}
where the values of the coefficients $A_j, B_j$ were chosen randomly and uniformly from the interval $[-6,6]$. Then 4 discontinuity points were added to the continuous functions. The coordinates of these points $\omega_1, \; \omega_2, \; \omega_3, \; \omega_4$ were set randomly and uniformly from the ranges $[0,0.25), \; [0.25,0.5), \; [0.5,0.75), \; [0.75,1]$ respectively. The jump values $\delta_1, \; \delta_2, \; \delta_3, \; \delta_4$ at discontinuities were also randomly selected from the range $[-50,50]$. These operations produced a discontinuous function of the form
\begin{equation}\label{seriaFunction_jump}
\varphi(x) =  f(x)+
\begin{cases}
	\delta_1, &0 \leq x < \omega_1\\
	\delta_2, &\omega_1 \leq x < \omega_2\\
	\delta_3, &\omega_2 \leq x < \omega_3 \;, \;\;\; x\in[0,1].\\
	\delta_4, &\omega_3 \leq x < \omega_4\\
	\delta_1, &\omega_4 \leq x \leq 1.0
\end{cases}
\end{equation}

Since the solution of the test problem assumes that the global optimizer $x^*$ is known, its value was preemptively estimated for each function of the series by iterating through all nodes of a uniform grid with the sufficiently small step.

The first experiment was performed on a series of 1000 problems with continuous functions of the form (\ref{seriaFunction}). Table~\ref{tab1} shows the number of problems solved, the average number of iterations performed by different methods, and the number of trials performed by the methods during the search. The problem was considered solved if the condition $|x^k-x^* | \leq \delta$ was satisfied for the trial point $x^k$, where $x^*$ is the known solution of the problem and $\delta= 10^{-2}$. At the same time, the accuracy used in the termination criteria for all methods was $\epsilon = 10^{-3}$.
